Hippocampal segmentation with MAGeT

### 2013-02-03 paper.Rnw
Outline of the intro

% atlas-based methods -> a prior feature based methods?
%
% - basic multi-atlas method, 
%  Collins2010 - .. seg of HC and AM by augmenting ANIMAL
%    - multi-atlas 80 atlases, atlas selection (NMI)
%  Aljabar2009 - atlas selection and effect on accuracy
% - extensions and practical applications
%  Leung2010 - X-sectional and longitudinal HC vol in MCI, AD
%    - (MAPS), multi-atlas + STAPLE. 55 atlases. baseline+1yr of ADNI seg.
%    - atrophy rates
%  Lotjonen2010 - Fast and robust
%    - IBSR/ADNI.  
% - Coutin's use of shape-based priors
%
% - multi-atlas with specialised/small atlas libraries
%   - Fischl - freesurfer, high-resolution atlases + probability priors
%   - Yushkevitch - invivo focal HF-MRI 
%  
%  Heckemann2011 -  Automatic morphometry in AD and MCI 
%    - (MAPER). multi-atlas. 30 atlases, entire ADNI screening  
%  Mouiha2011b - Hippocampal atrophy rates in AD
%    - compare SNT vs FreeSurfer on ADNI
%
%
Fully-automated segmentation techniques require no human expertise beyond that
of an initial configuration.  Multi-atlas automated segmentation algorithms make
use of a series of expertly segmented MRIs (the atlas library) as models to
guide the segmentation of a query subject's MRI, and achieve some of the best
automated hippocampal segmentation accuracies to-date

\citep{Aljabar2009,Collins2010,Heckemann2011,Leung2010,Lotjonen2010,Leung2010,Wolz2010}.
The basic strategy is as follows. Each atlas is warped to fit a subject's
neuroanatomy using nonlinear registration techniques
\citep{Collins1995,Klein2009}.  Atlas labels are then transformed by this
warping to the subject's space, and a {\it label fusion} technique, such as
voxel-wise voting, is used to merge the competing label definitions from each
atlas into a final segmentation for the subject.

>>>>

Fully-automated segmentation techniques require no human expertise beyond that
of an initial configuration.  Multi-atlas automated segmentation algorithms make
use of a series of expertly segmented MRIs (the atlas library) as models to
guide the segmentation of a query subject's MRI, and achieve some of the best
automated hippocampal segmentation accuracies to-date

The basic strategy is as follows. Each atlas is warped to fit a subject's
neuroanatomy using nonlinear registration techniques
\citep{Collins1995,Klein2009}.  Atlas labels are then transformed by this
warping to the subject's space, and a {\it label fusion} technique, such as
voxel-wise voting, is used to merge the competing label definitions from each
atlas into a final segmentation for the subject.

\citep{Aljabar2009,Collins2010,Heckemann2011,Leung2010,Lotjonen2010,Leung2010,Wolz2010}.

## automated segmentation and the multi-atlas approach
- automated segmentation techniques require no human intervention
- model-based automatic segmentation approaches employ models of feature
  variations as constraints during segmentation. 
- multi-atlas segmentation techniques use many atlases and perform well
- basic multi-atlas methods were pioneered by Heckmann et al in the MAPER
  algorithm Heckemann 2006 and 2010. The strategy is as follows: 
- individual atlases are treated as classifiers, registration is used to map the
  classifiers to a subject space, and decision fusion is used to combine
  expertise from classisfiers. define: atlas library
- In Heckemann2006 a library of whole brain atlases are used, but Aljabar2009
  shows that selecting a subset of the atlases most similar (using
  cross-correlation of image intensities) to a subject and fusing only those
  labels leads to better performance.  define: atlas selection.
- Warfield2004: STAPLE as an alternative decision fusion technique?

## application to  hippocampus
- There are many applications of multi-atlas techniques to hippocampal
  segmentation
- collins2010 apply the multi-atlas approach to hippocampal and amygdala
  segmentaiton using an atlas library of 80 atlases segmented using a custom
  protocol (images selected from the ICBM152 dataset), the ANIMAL nonlinear
  registration algorithm, normalised mutual information for atlas selection, and
  majority vote for decision fusion, achieves a Dice Kappa similarity of 0.886.
- Leung2010 (MAPS) uses an atlas library of 55 atlases (segmented using a
  custom protocol) and STAPLE decision fusion to segment ADNI1 baseline and 12
  month dataset. After selecting parameters for the best performance on the ADNI
  dataset, they were able to achieve a Dice Kappa score of 0.89 compared to the
  manual segmentations supplied with the ADNI data (see Methods for more
  information on the ADNI dataset). 
- Heckmann2011 used the MAPER algorithm (described earlier) with 30 atlases  to
  segment ADNI1 1.5 and 3T baseline images, achieving a mean Dice Kappa
  similarity score of 0.889. 
- Lotjonen2012 explored various atlas selection methods, and used a custom
  non-linear registration method based on intensity differences, with
  post-processing of fused segmentations taking into account intensity
  differences to tidy up the segmentations.  They applied this to 60 subjects
  taken from the ADNI1 baseline dataset, using an atlas library of 30 subjects
  (also taken from the ADNI1 baseline dataset) and achieved an mean Dice Kappa
  similarity score of 0.885
- Wolz2010 - differ significantly from the basic multi-atlas strategy in that
  the atlas library is grown, starting with a set of manually labelled atlases
  and successively incorporating unlabelled target images after labelling.
  The sequence in which target images are labelled is such that the similarity
  between the atlas images and the target images is minimised at each step,
  effectively allowing for deformations between very dissimilar images to be
  broken up into sequences of smaller deformations.  With an atlas library of
  30 MR images, LEAP was used to segment the ADNI1 baseline dataset, achieving a
  mean Dice score of 0.85 with manual segmentations.

## specialised automated segmentation techiques of note
- Fischl 2002 - describe a method which uses a probabilistic atlas of anatomical
  and tissue classes along with spatial constraints for class labels encoded
  using a Markov random field model. This forms the basis of FreeSurfer's whole
  brain segmentation algorithm.
- Fischl 2009 use a probabilistic atlas of anatomical classes as well as a
  likelihood model of how those classes translate into MR image intensities.
  Trained on manual segmentations of high resolution MR images, Bayesian
  inference is used to classify voxels in an unseen image.  This forms the basis
  of the FreeSurfer hippocampal subfield segmentation algorithm. 
- Yushkevitch2010 describe a semi-automated method for hippocampal subfield
  segmentation of focal T2 images.  The unlabelled MR image must be manually
  partitioned into 'head', 'body' and 'tail', and then multi-atlas methods are
  used to segment the image (ANTS registration, and similarity-weighted vote).
  Finally, an AdaBoost-based bias correction classifier trained on texture,
  spatial location and segmentation is applied to fix mislabelled voxels.

## with all of these different methods, how much does the atlas library matter?
- all of these methods require a sizeable atlas library which, as we've
  discussed, requires significant effort to create. how much does an atlas
  library matter for the performance of an automated method?
- As noted, Aljabar2009's work on atlas selection shows clearly that the
  composition of the atlas library has a significant effect on accuracy of
  automated segmentation.
- what anatomical features play a role here? How does protocol impact?  
- Carmichael2005 - explores which aspects of the single-atlas strategy matter
  most: registration, similarity between atlas and gold-standard delineation
  protocol, and neuroanatomical similarity between the atlas image and subject
  image. They found that most flexible registration (non-linear) method, atlas
  segmentation that follows the same protocol as the gold-standard and,
  importantly for us, an atlas with neuroanatomical features that match the
  subject image lead to the highest overlap of labels. 
- Nestor2012 explored the effect of segmentation protocol on automated
  multi-atlas segmentation performance of the hippocampus in the ADNI1 1 Year
  dataset, and found that protocols that more inclusive protocols which include
  dorsal white-matter and posterior anatomy tended to produce higher overlap and
  better accuracy at distinguishing disease classes.

## the problem of needing a large atlas library, and the benefit of having a
   small one
- time expensive: if you have one atlas from a difficult to obtain source?
  e.g. histological atlas
- customising definitions is tough, but necessary to ask research questions

%  Chupin's use of shape-based priors 
%   - Mouiha2011b - Hippocampal atrophy rates in AD: compare SNT vs FreeSurfer
%     on ADNI
%

## introducing MAGeT brain for bootstrapping automated hippocampal segmentation
- MAGeT bootstrapping: tune the template library to your dataset, reduce the
  number of input atlases you need, with the upshot that with fewer atlases it's
  that much easier to change definitions in order to test hypothesis
